---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

First, we need to load in required packages.

```{r packages, message=FALSE}
library(dplyr)
library(lubridate)
```


## Loading and preprocessing the data

Now, we read in the data.  

```{r read_in_data}
movement_data <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?

The first question we must answer is "What is the mean total number of steps taken per day?".

In order generate this first figure, we need to group the data by date.  Next, for each unique date, we total the steps taken.

```{r total_steps}
total_steps <- movement_data %>% group_by(date) %>% summarize(sum_steps = sum(steps, na.rm = T))%>% mutate(date = ymd(date)) 

head(total_steps)
```

Then, we generate a histogram.

```{r total_steps_figure}
hist(total_steps$sum_steps, breaks = 15, ylab = "Frequency", xlab = "Total Steps", main = "Total Number of Steps per Day", col = "white")
abline(v = mean(total_steps$sum_steps), col = "red", lwd = 2)
abline(v = median(total_steps$sum_steps), col = "blue", lwd = 2)
```

A quick look at this histogram will tell us the mean (red) total number of steps taken per day over this 2 month period.  It will also give us information on the median (blue) and the mode (highest bar).

Here, we see that the mean total number of steps falls short of 10,000 steps a day, while the median is slightly higher than 10,000 steps a day.  This can be shown computationally, as well.

```{r summary_calculations}
summary(total_steps$sum_steps)
```

This validates what our graph showed us:  the median total steps per day is 10,395 steps.  The mean is 9354 steps.


## What is the average daily activity pattern?

The next question we are tasked with answering is "What is the average daily activity pattern?".  This we must answer with a time series plot.

For this, we need to group the data by the interval variable, then take the average number of steps.

```{r steps_interval_figure}
interval_steps <- movement_data %>% group_by(interval) %>% summarize(mean_steps = mean(steps, na.rm = T))

plot(interval_steps, type = "l", ylab = "Average Number of Steps", xlab = "5-Minute Interval", main = "Average Daily Activity")
```

Next, we can look at exactly which interval had the highest number of steps, on average.

```{r }
highest_interval <- which.max(interval_steps$mean_steps) %>% interval_steps[.,]

print(highest_interval)
```

Here, we see that interval 835 has the highest average, about 206 steps.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
