Quantified Self Movement
========================

##Getting started: Loading Packages and Figures

First, load in required packages.

```{r packages, message=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

Read in the data.  

```{r read_in_data}
movement_data <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE)
```

##What is the mean total number of steps taken per day?

To generate this first figure, the data need to be grouped by date.  Then, all the steps need to be totalled for each day.

```{r total_steps}
total_steps <- movement_data %>% group_by(date) %>% summarize(sum_steps = sum(steps, na.rm = T))%>% mutate(date = ymd(date)) 

head(total_steps)
```

A quick look at this histogram will tell us the mean (red)total number of steps taken per day over this 2 month period.  It will also give us information on the median (blue) and the mode (highest bar).

```{r total_steps_figure}
hist(total_steps$sum_steps, breaks = 15, ylab = "Frequency", xlab = "Total Steps", main = "Total Number of Steps per Day", col = "white")
abline(v = mean(total_steps$sum_steps), col = "red", lwd = 2)
abline(v = median(total_steps$sum_steps), col = "blue", lwd = 2)
```

Here, we see that the mean total number of steps falls short of 10,000 steps a day, while the median is slightly higher than 10,000 steps a day.  This can be shown computationally, as well:

```{r summary_calculations}
summary(total_steps)
```